---
title: "Programatically Create Formulas in R"
categories: [R, package-dev]
date: 2023-01-17
---

I'm working on a prototype of a tool using R to create some regressions. I'd like to take in independent and dependent variables as character vectors and use those to create a formula. It took me a minute to figure out how to programatically create a formula. It's a bit tricky because formula require unquoted objects (symbols).

The trick is to use the function `reformulate()` ([thanks StackOverflow](https://stackoverflow.com/questions/12967797/is-there-a-better-alternative-than-string-manipulation-to-programmatically-build)).

The syntax is `reformulate(x_vars, y_var)`

```{r}
form <- reformulate(c("x1", "x2"), "y")
form
```

Nice. Throw it into a function.

```{r}
make_lm <- function(.data, y, x) {
  form <- reformulate(x, y)
  lm(form, .data)
}
```


Now try it out :) 

```{r}
make_lm(
  iris, 
  y = "Petal.Width",
  x = c("Sepal.Length", "Petal.Length")
)
```


This can be a pretty hand pattern in package development. I hope it helps you. 
