---
title: Evaluate strings as code 
date: "2024-06-13"
categories: [r]
---

Prompted by a post on Mastodon, I wanted to explore how to evaluate an R string as code.

<iframe src="https://mastodon.cloud/@nxskok/112610810465402574/embed" width="400" allowfullscreen="allowfullscreen" sandbox="allow-scripts allow-same-origin allow-popups allow-popups-to-escape-sandbox allow-forms"></iframe>

This is generally a pretty common pattern that I have myself encountered in the past and had to work through a solution for—many times. 

## The Problem

How can I programatically create and execute valid R code? 

## A solution

In this case, the problem space is quite simple: 

1. given a package name and 
2. a dataset name
3. extract the dataset as an object

You can typically extract datasets from a package's namespace. This looks like `{pkgname}::{dataset}`. 

We can create this string simply like so: 

```{r}
pkg <- "dplyr"
dataset <- "starwars"
dataset_str <- paste0(pkg, "::", dataset)
```

### Evaluating R code 

Then, we need to be able to evaluate this code. I find [`{rlang}`](https://rlang.r-lib.org/reference/) to be very handy. 

To convert a string into an expression, use `rlang::parse_expr()`


```{r}
library(rlang)
to_eval <- parse_expr(dataset_str)
to_eval
```

This creates a `language` type object. 

We can now pass this into `rlang::eval_bare()` to evaluate the string and run the R code and store the result into an R object.

```{r}
result <- rlang::eval_bare(to_eval)
result
```


## Alternative solution

Here is an alternative solution which uses the `data()` function. Then, assuming the name of the dataset is created in the environment, fetches it using `get()`.

```{r}
englue("data({dataset}, package = '{pkg}')") |>
  parse_expr() |>
  eval_bare()

res <- get(dataset)
res
```

There are issues with this in that you can also end up overwriting things. We can create a new environment if we'd like as well. 


```{r}
# create a custom environment to store stuff
my_env <- rlang::env()

englue("data({dataset}, package = '{pkg}')") |>
  parse_expr() |>
  eval_bare(my_env)

# get it from the environment
res <- get(dataset, envir = my_env)
res
```