---
title: "Create and edit TOML in R with {tomledit}"
date: 2025-04-10
author: Josiah Parry
categories: [r,  toml, rust, extendr]
---

[`{tomledit}`](https://extendr.github.io/tomledit/) v0.1.1 has found its way onto [CRAN](https://cran.r-project.org/web/packages/tomledit/index.html). 


`{tomledit}` is a package for _creating_ and _editing_ TOML files from R with support for reading as well. 


The most basic use of `tomledit` is via `toml()`.  `toml()` creates a `Toml` object from named arguments passed to `...`. 

```{r}
library(tomledit)

toml(person = list(age = 30L, name = "Wilma"))
```


## v0.1.1 Features

This newest release supports the use of arrays with inline tables. This feature comes as a request from [@dpastoor](https://github.com/dpastoor) to support the experimental `rproject.toml` file for [`rv`](https://github.com/A2-ai/rv).

:::{.aside}
I'm bullish on `rv` as a new alternative to `{renv}`. I think it will be a great addition to the R community.
:::

This new feature allows us to have a list of unnamed lists inside of our TOML. 

Below we create an item called `repositories` which is an array of inline tables containing the alias and url to a CRAN-like repository. 

Similarly, the `dependencies` item is an array of _both_ inline-tables and strings. This new feature adds more flexibility to the type of TOML that we can create. 

```{r}
r_proj_toml <- list(
  name = "upgrade", 
  r_version = "4.4",
  repositories = list(
    list(alias = "gh-pkg-mirror", url = "https://a2-ai.github.io/gh-pkg-mirror/2024-02-22"),
    list(alias = "RSPM", url = "https://packagemanager.posit.co/cran/2024-02-22"),
    list(alias = "new-mirror", url = "https://a2-ai.github.io/gh-pkg-mirror/2024-12-04"),
    list(alias = "new-rspm", url = "https://packagemanager.posit.co/cran/2024-12-04")
  ),
  dependencies = list(
    list(name = "pmplots", repository = "new-mirror"),
    "pmtables",
    "bbr",
    list(name = "ggplot2", repository = "new-rspm")
  )
) 

as_toml(list(project = r_proj_toml))
```