---
title: Complete spatial randomness
date: '2022-11-08'
categories: [spatial, rstats, sfdep]
tags: [rspatial]
---

< this is a cliche about Tobler's fist law and things being related
in space>. Because of Tobler's first law, spatial data tend to not follow any
specific distribution. So, p-values are sort of...not all that accurate most of the time. 
P-values in spatial statistics often take a "non-parametric" approach instead
of an "analytical" one. 

Consider the t-test. T-tests make the assumption that data are 
coming from a normal distribution. Then p-values are derived from the 
cumulative distribution function. The alternative hypothesis, then, is 
that the true difference in means is not 0.  

In the spatial case, our alternative hypothesis is generally "the observed
statistic different than what we would expect under complete spatial
randomness?" But what really does that mean? To know, we have to 
simulate spatial randomness. 

There are two approaches to simulating spatial randomness that I'll go
over. One is better than the other. First, I'm going to describe the 
less good one: bootstrap sampling.

Load the super duper cool packages. We create queen contiguity neighbors and row-standardized weights.


```{r, message = FALSE}
library(sf)
library(sfdep)
library(tidyverse)

grid <- st_make_grid(cellsize = c(1, 1), n = 12, offset = c(0, 0)) |> 
  as_tibble() |> 
  st_as_sf() |> 
  mutate(
    id = row_number(),
    nb = st_contiguity(geometry),
    wt = st_weights(nb)
    )
```

Let's generate some spatially autocorrelated data. This function is a little slow, but it works.

```{r, message = FALSE}
nb <- grid[["nb"]]
wt <- grid[["wt"]]

x <-  geostan::sim_sar(w = wt_as_matrix(nb, wt), rho = 0.78)
```

## Bootstrap sampling

Under the bootstrap approach we are sampling from existing spatial 
configurations. In our case there are 144 existing neighborhoods.
For our simulations, we will randomly sample from existing neighborhoods
and then recalculate our statistic. It helps us by imposing randomness
into our statistic. We can then repeat the process `nsim` times. There is 
a limitation, however. It is that there are only `n - 1` 
possible neighborhood configurations per location.

Here we visualize a random point and it's neighbors.

```{r }

# for a given location create vector indicating position of
# neighbors and self
color_block <- function(i, nb) {
  res <- ifelse(1:length(nb) %in% nb[[i]], "neighbors", NA)
  res[i] <- "self"
  res
}
```

Plot a point and its neighbors

```{r }
grid |> 
  mutate(block = color_block(sample(1:n(), 1), nb)) |> 
  ggplot(aes(fill = block)) +
  geom_sf() +
  labs(title = "Point and it's neighbors")
```

For bootstrap we grab a point and then the neighbors from another point. 
This function will randomize a `nb` list object.

```{r }
color_sample_block <- function(i, nb) {
  index <- 1:length(nb)
  not_i <- index[-i]
  
  sample_block_focal <- sample(not_i, 1)
  
  res <- rep(NA, length(index))
  
  res[nb[[sample_block_focal]]] <- "neighbors"
  res[i] <- "self"
  res
}

# visualize it
grid |> 
  mutate(block = color_sample_block(sample(1:n(), 1), nb)) |> 
  ggplot(aes(fill = block)) +
  geom_sf() +
  labs(title = "Point and random point's neighbors")
```

Often, we will want to create a reference distribution by creating a large
number of simulations—typically 999. As the simulations increase in size, we are limited in the amount of 
samples we can draw. The number of neighborhoods becomes limiting! 

Say we want to look at income distribution in Boston and the only data we 
have is at the census tract level. I happen to know that Boston has 207 
tracts. If we want to do 999 simulations, after the 206th simulation, we 
will likely have gone through all over the neighborhood configurations! 

How can we do this sampling? For each observation, we can sample another
location, grab their neighbors, and assign them as the observed location's
neighbors. 

## Bootstrap simulations

In sfdep, we use spdep's `nb` object. These are lists that store the row
position of the neighbors as integer vectors at each element. 

> If you want to learn more about neighbors [I gave a talk](https://youtu.be/i_MA1U6SJ1Y?t=1301) at NY Hackr MeetUp a few months ago that might help.

Here I define a function that samples from the positions (`index`), 
then uses that sample to shuffle up the existing neighborhoods and 
return a shuffled `nb` object. Note that I add the `nb` class back
to the list. 


```{r }
bootstrap_nbs <- function(nb) {
  # create index
  index <- 1:length(nb)
  # create a resampled index
  resampled_index <- sample(index, replace = TRUE)
  # shuffle the neighbors and reassign class
  structure(nb[resampled_index], class = c("nb", "list"))
}
```

Let's compare some observations

```{r }
nb[1:3]

bootstrap_nbs(nb)[1:3]
```

Here we can see the random pattern. Look's like there is fair amount of 
clustering of like values. 

```{r }
grid |> 
  mutate(x = classInt::classify_intervals(x, 7)) |> 
  ggplot(aes(fill = x)) +
  geom_sf(color = NA, lwd = 0) +
  scale_fill_brewer(type = "div", palette = 5, direction = -1) +
  theme_void() 
```

With the weights and the neighbors we can calculate the global Moran.
I'll refer to this as the "observed." Store it into an object called 
`obs`. We'll need this to calculate a simulated p-value later.

```{r }
obs <- global_moran(x, nb, wt)
obs[["I"]]
```

`r round(obs[["I"]], 2)` is a fair amount of positive spatial autocorrelation indicating 
that like values tend to cluster. But is this due to random chance, or
does it depend on where these locations are?
Now that we have the observed value of Moran's I, we can simulate the 
value under spatial randomness using the bootstrapped sampling. To do so,
we bootstrap sample our neighbors, recalculate the weights and then 
the global Moran.
Now, if you've read my [vignette on conditional permutation](https://sfdep.josiahparry.com/articles/conditional-permutation.html), you know what
is coming next. We need to create a reference distribution of the global
Moran under spatial randomness. To do that, we apply our boot strap `nsim`
times and recalculate the global Moran with each new neighbor list. I love
the function `replicate()` for these purposes. 

```{r }
nsim = 499 
```

> _Also, a thing I've started doing is assigning scalars / constants with an equals sign because they typically end up becoming function arguments._

```{r }
reps <- replicate(
  nsim, {
    nb_sim <- bootstrap_nbs(nb)
    wt_sim <- st_weights(nb_sim)
    global_moran(x, nb_sim, wt_sim)[["I"]]
  }
)

hist(reps, xlim = c(min(reps), obs[["I"]]))
abline(v = obs[["I"]], lty = 2)
```

### Bootstrap limitations

That's all well and good, but let's look at this a bit more. Since we're
using the bootstrap approach, we're limited in the number of unique 
combinations that are possible. Let's try something. Let's calculate
the spatial lag `nsim` times and find the number of unique values that
we get. 


```{r }
lags <- replicate(
  nsim, {
    # resample the neighbors list
    nb_sim <- bootstrap_nbs(nb)
    # recalculate the weights
    wt_sim <- st_weights(nb_sim)
    # calculate the lag
    st_lag(x, nb_sim, wt_sim)
  }
)

# cast from matrix to vector
lags_vec <- as.numeric(lags)

# how many are there?
length(lags_vec)

# how many unique?
length(unique(lags_vec))
```

See this? There are only 144 unique value! That isn't much! Don't believe me? Run `table(lags_vec)`.
For each location there are only a limited number of combinations that
can occur. 

## Conditional Permutation

Now, here is where I want to introduce what I view to be the superior
alternative: conditional permutation. Conditional permutation was 
described by Luc Anselin in his seminal 1995 paper. The idea is that we
hold an observation constant, then we randomly assign neighbors. This 
is like the bootstrap approach but instead of grabbing a random 
observation's neighborhood we create a totally new one. We do this be
assigning the neighbors randomly from all possible locations.

Let's look at how we can program this. For each location we need to
sample from an index that excludes the observation's position. Further
we need to ensure that there are the same number of neighbors in each
location (cardinality).

```{r }
permute_nb <- function(nb) {
  # first get the cardinality
  cards <- st_cardinalties(nb)
  # instantiate empty list to fill
  nb_perm <- vector(mode = "list", length = length(nb))
  
  # instantiate an index
  index <- seq_along(nb)
  
  # iterate through and full nb_perm
  for (i in index) {
    # remove i from the index, then sample and assign
    nb_perm[[i]] <- sample(index[-i], cards[i])
  }
  
  structure(nb_perm, class = "nb")
}


nb[1:3]
permute_nb(nb)[1:3]
```

Now, let's repeat the same exercise using conditional permutation. 

```{r }
lags2 <- replicate(
  nsim, {
    nb_perm <- permute_nb(nb)
    st_lag(x, nb_perm, st_weights(nb_perm))
  }
)

lags2_vec <- as.numeric(lags2)
  
length(unique(lags2_vec))
```

There are farrrrr more unique values. In fact, there is a unique value
for each simulation - location pair. If we look at the histograms, 
the difference is even more stark. The conditional permutation approach actually begins to represent a real distribution.


```{r }
par(mfrow = c(1, 2))
hist(lags_vec, breaks = 20) 
hist(lags2_vec, breaks = 20)
```

So, this is all for me to say that bootstrapping isn't it for creating
simulated distributions for which to calculate your p-values. 
