---
title: "Read a CSV in a production API"
subtitle: "{plumber} and multipart request #RinProd"
date: "2024-06-25"
categories: [plumber, r, prod]
---

Deploying RESTful APIs is the way to put any language into production. R is not any different. 

One challenge when making APIs is handling files.

Uploading files is done typically with a [_multipart request_](https://swagger.io/docs/specification/describing-request-body/multipart-requests/).  

> "[they] combine one or more sets of data into a single body…. You typically use these requests for file uploads and for transferring data of several types in a single request (for example, a file along with a JSON object)."

:::{.aside}
[source: Swagger doc](https://swagger.io/docs/specification/describing-request-body/multipart-requests/)
:::

## Handling multipart requests in R

You can process them using the [`{mime}`](https://github.com/yihui/mime) package.

:::{.aside}
Named after ["mime types"](https://en.wikipedia.org/wiki/Media_type) not [Mr. Mime](https://www.pokemon.com/us/pokedex/mr-mime)
:::

[`{plumber}`](https://rplumber.io) provides access to the body of a request using the `req` argument. 

```r
#* @post /upload
upload <- function(req, res) {
    # body
}
```

To access the structure of request use [`mime::parse_multipart(req)`](https://rdrr.io/github/yihui/mime/man/parse_multipart.html). 

Modifying the function like so will return json from the API

```r
#* @post /upload
upload <- function(req, res) {
    mp <- mime::parse_multipart(req)
    mp
}
```

**Save this as plumber.R**

### Run your API

In your terminal (from the same working directory as `plumber.R`) run `R -e 'plumber::plumb("plumber.R")$run(port = 3000)'`

This will give you a background API to call. 

## Making a multipart request

Use [`httr2`](https://httr2.r-lib.org) to create the multipart request. 

- Start the request with `request()`
- Base the request object to `req_body_multipart()` to add data 
- Use key-value pairs to `req_body_multipart(...)` to add data
    - Note that values must be a string so create the json yourself
- Send the request using `req_perform()`


Here we give it a unique ID and add a sample of data 

```{r}
library(httr2)

resp <- request("http://127.0.0.1:3000/upload") |>
  req_body_multipart(
    id = ulid::ulid(),
    sample = jsonify::to_json(sample(1:100, 10), unbox = TRUE)
  ) |>
  req_perform()

resp
```

We extract the data using `resp_body_string()` and process it using 
```{r}
resp_body_string(resp) |>
  RcppSimdJson::fparse()
```

## Adding files

We'll create a tempory file containing the `iris` data.frame and send this to the API endpoint.

These two lines: 

1. Create a temporary csv file
2. Write the data frame to the temporary file 

:::{.callout-tip}
This is a very handy trick that you might be able to adapt to many other circumstances. Temporary files are _very_ useful. 
:::

```{r}
tmp <- tempfile(fileext = ".csv")
readr::write_csv(head(iris), tmp)
```


Next we need to upload the file to our request. Do this using `curl::form_file()`. You need to provide a path to the file. In this case, it will be the temporary file. 

```{r}
resp <- request("http://127.0.0.1:3000/upload") |>
  req_body_multipart(
    file = curl::form_file(tmp)
  ) |>
  req_perform()

resp_body_string(resp) |>
  jsonify::pretty_json()
```

In this case `file` is a named list. `mime` stores the file in a temporary path accessible via `datapath`. So let's try adding an API endpoint to read a csv file. 


## Read CSV in Plumber API


Here we read the csv from the path. We would probably need to add some better checks here. Like checking that the field actually exists in `mp` but the error will be propagates as a 500 status anyways. 

Something is always better than nothing. Just like this blog post. 

```r
#* @post /read_csv
function(req, res) {
  mp <- mime::parse_multipart(req)
  readr::read_csv(mp$file$datapath)
}
```


## Send CSV to API

Here is how we can send the csv to the API

```{r}
resp <- request("http://127.0.0.1:3000/read_csv") |>
  req_body_multipart(
    file = curl::form_file(tmp)
  ) |>
  req_perform()

resp_body_string(resp) |>
  jsonify::pretty_json()
```

Note that the response is just nice json.

We can parse that back doing a full round trip:


```{r}
resp_body_string(resp) |>
  RcppSimdJson::fparse()
```


# Whole API:


```{r, filename="plumber.R", eval = FALSE}
library(plumber)
# R -e 'plumber::plumb("plumber.R")$run(port = 3000)'
#* @post /upload
upload <- function(req, res) {
  mp <- mime::parse_multipart(req)
  mp
}

#* @post /read_csv
function(req, res) {
  mp <- mime::parse_multipart(req)
  readr::read_csv(mp$file$datapath)
}

```

##  Scale your APIs

Use [`Valve`](https://valve.josiahparry.com/) to scale and deploy your applications to production. 

It kicks ass tbh. 